Analysis for Project

### 2021-11-15 analysis/analysis.Rmd
Clean up analysis.Rmd
I put names on all code chunks, placed tabular output into proper kable tables and  wrote some text for the front end.

# Grade Retention Profile

@@

```{r tab-grade-by-age}
table(acs$age, acs$current_grade)
```

@@

```{r}
tapply(acs$below_exp_grade,acs[,c("age","current_grade")], mean)
```

@@

```{r}
round(tapply(acs$below_exp_grade, acs$current_grade, mean), 3)*100
```

@@

```{r}
round(tapply(acs$below_exp_grade, acs[,c("current_grade","year")], mean), 3)*100
```

Some evidence of declining grade retention prevalence at higher grades.

# Lets look at the multiracial sample

How big is it?

```{r}
acs %>%
  filter(str_detect(race, "/") & !str_detect(race, "Multiracial")) %>%
  ggplot(aes(x=reorder(race, race, length))) +
  geom_bar()+
  coord_flip()+
  labs(x=NULL)+
  theme_bw()
```

```{r}
resources_race <- acs %>% 
  group_by(race) %>%
  summarize(prop_held_back=mean(below_exp_grade),
            mean_age=mean(age),
            prop_foreign_born=mean(foreign_born),
            mean_fam_inc=mean(family_income),
            prop_own_home=mean(own_home),
            prop_hs_diploma=mean(degree_father>="HS diploma" | 
                                   degree_mother>="HS diploma"),
            prop_ba_degree=mean(degree_father>="BA degree" | 
                                  degree_mother>="BA degree"),
            prop_parents_married=mean(parents_married),
            mean_age_parents=mean((age_birth_mother+age_birth_father)/2),
            prop_parent_foreign_born=mean(foreign_born_father | 
                                            foreign_born_mother)) %>%
  ungroup() %>%
  filter(!str_detect(race,"Multiracial")) %>%
  mutate(multiracial=str_detect(race, "/"))
```

```{r}
ggplot(resources_race, aes(x=reorder(race, mean_fam_inc, mean), y=mean_fam_inc,
                           fill=multiracial))+
  geom_col()+
  coord_flip()+
  labs(x=NULL, y="mean family income")
```

```{r}
ggplot(resources_race, aes(x=reorder(race, prop_own_home, mean), y=prop_own_home,
                           fill=multiracial))+
  geom_col()+
  coord_flip()+
  labs(x=NULL, y="Proportion who own their own home")
```

```{r}
ggplot(resources_race, aes(x=reorder(race, prop_hs_diploma, mean), y=prop_hs_diploma,
                           fill=multiracial))+
  geom_col()+
  coord_flip()+
  labs(x=NULL, y="Proportion where one parent has HS diploma")
```

```{r}
ggplot(resources_race, aes(x=reorder(race, prop_ba_degree, mean), y=prop_ba_degree,
                           fill=multiracial))+
  geom_col()+
  coord_flip()+
  labs(x=NULL, y="Proportion where one parent has college degree")
```

## Raw differences in grade retention

```{r}
ggplot(resources_race, aes(x=reorder(race, prop_held_back, mean), y=prop_held_back,
                           fill=multiracial))+
  geom_col()+
  coord_flip()
```

## Build some models

@@

```{r}
model0 <- glm(below_exp_grade~race, data=design_matrix, family=binomial)
model1 <- update(model0, 
                 .~.+y2011+y2012+y2013+y2014+y2015+y2016+y2017+y2018+y2019+
                   current_grade1st+current_grade2nd+current_grade3rd+
                   current_grade4th+current_grade5th+current_grade6th+
                   current_grade7th+current_grade8th+current_grade9th+
                   current_grade10th+current_grade11th+current_grade10th+
                   current_grade11th+current_grade12th+
                   met_non_metro+met_central_city+met_non_central_city+
                   met_city_indeter+
                   s1+s2+s3+s4+s5+s6+s7+s8+s9+s10+s11+s12+s13+s14+s15+s16+s17+
                   s18+s19+s20+s21+s22+s23+s24+s25+s26+s27+s28+s29+s30+s31+s32+
                   s33+s34+s35+s36+s37+s38+s39+s40+s41+s42+s43+s44+s45+s46+s47+
                   s48+s49+s50)
model2 <- update(model1, .~.+foreign_born+foreign_born_mother+
                   foreign_born_father)
model3 <- update(model2, .~.+degree_motherHS+degree_motherAA+degree_motherBA+
                   degree_motherG+degree_fatherHS+degree_fatherAA+
                   degree_fatherBA+degree_fatherG)
model4 <- update(model3, .~.+income_trans+own_home+parents_married)

@@

```{r}
#dateset for predict
df_pred <- data.frame(race=levels(acs$race),
                      y2011=0, y2012=0, y2013=0, y2014=0, y2015=0, y2016=0,
                      y2017=0, y2018=0, y2019=0,
                      current_grade1st=0, current_grade2nd=0, current_grade3rd=0,
                      current_grade4th=0, current_grade5th=0, current_grade6th=0,
                      current_grade7th=0, current_grade8th=0, current_grade9th=0,
                      current_grade10th=0, current_grade11th=0, current_grade12th=0,
                      met_non_metro=0, met_central_city=0, met_non_central_city=0,
                      met_city_indeter=0,
                      s1=0, s2=0, s3=0, s4=0, s5=0, s6=0, s7=0, s8=0, s9=0, s10=0,
                      s11=0, s12=0, s13=0, s14=0, s15=0, s16=0, s17=0, s18=0, s19=0,
                      s20=0, s21=0, s22=0, s23=0, s24=0, s25=0, s26=0, s27=0, s28=0,
                      s29=0, s30=0, s31=0, s32=0, s33=0, s34=0, s35=0, s36=0, s37=0,
                      s38=0, s39=0, s40=0, s41=0, s42=0, s43=0, s44=0, s45=0, s46=0,
                      s47=0, s48=0, s49=0, s50=0,
                      degree_motherHS=0, degree_motherAA=0, degree_motherBA=0,
                      degree_motherG=0, degree_fatherHS=0, degree_fatherAA=0, 
                      degree_fatherBA=0, degree_fatherG=0,
                      foreign_born=0, foreign_born_father=0, foreign_born_mother=0,
                      income_trans=0, own_home=0, parents_married=0)
```

@@

```{r}
coef_table <- calculate_cond_means(model0)
```

@@

```{r}
g <- coef_table %>%
  filter(mrace=="Black/White") %>%
  ggplot(aes(x=term, y=estimate, color=multiracial,
             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial & mrace=="Black/White"))+
  geom_point()+
  coord_flip()+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="predicted probability of being behind expected grade",
       title="Probability of being behind grade for biracial black/white respondents and\nmonoracial comparison groups",
       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")

g+annotate("text", 1, 0.037, hjust="left",
             label="Grey bands give 83%\nconfidence interval around\nmonoracial estimates")+
  annotate("text", 2, 0.0475, hjust="left",
             label="Red band gives 83%\nconfidence interval around\nmidpoint estimate")+
  annotate("text", 2.5, 0.038, hjust="left",
             label="Actual biracial estimate\nshown with 83% confidence\ninterval")
```
This result is from the baseline model with no control variables, so this would be identical to the result you would get by just calculating raw proportions and their binomial standard errors. The grey bands for the monoracial groups bookmark the range of results we might expect and the red band gives what would be the midpoint value between the groups, adjusting for statistical uncertainty. I can then see where the actual biracial estimate falls within this range. The results here indicate that biracial black/white students have probabilities of grade retention much closer to the lower values of white than black students. However, their probability is still a bit higher than that for whites.

@@

```{r}
coef_table <- calculate_cond_means(model4)

ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial))+
  geom_point()+
  coord_flip()+
  facet_wrap(~mrace, ncol=3, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="predicted probability of being behind expected grade",
       title="Probability of being behind grade for biracial respondents and\nmonoracial comparison groups",
       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
```

@@

```{r, fig.height=12}
coef_table <- calculate_cond_means(model0) %>%
  mutate(model="crude")
coef_table <- calculate_cond_means(model1) %>%
  mutate(model="fixed effects") %>%
  bind_rows(coef_table)
coef_table <- calculate_cond_means(model2) %>%
  mutate(model="cultural resources") %>%
  bind_rows(coef_table)
coef_table <- calculate_cond_means(model4) %>%
  mutate(model="parent resources") %>%
  bind_rows(coef_table) %>%
  mutate(model=factor(model, levels=c("crude","fixed effects",
                                      "cultural resources",
                                      "parent resources")),
         mrace=factor(mrace, levels=c("Black/White",
                                      "Black/Latino",
                                      "Black/Asian",
                                      "Black/Indigenous",
                                      "White/Indigenous",
                                      "White/Latino",
                                      "White/Asian",
                                      "Latino/Asian",
                                      "Indigenous/Latino",
                                      "Indigenous/Asian")))

ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial))+
  geom_point()+
  coord_flip()+
  facet_grid(mrace~model, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="predicted probability of being behind expected grade",
       title="Probability of being behind grade for biracial respondents and\nmonoracial comparison groups",
       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
```

@@

## Sensitivity tests

```{r eval=FALSE}
#run separate models by grade groups
model4a <- update(model4, data=subset(acs, as.numeric(current_grade)<=6))
model4b <- update(model4, data=subset(acs, as.numeric(current_grade)>6 &
                                        as.numeric(current_grade)<=9))
model4c <- update(model4, data=subset(acs, as.numeric(current_grade)>=10))

>>>>

# Grade retention summary

@@

```{r tab-grade-by-age}
tbl <- table(acs$age, acs$current_grade)

tbl %>%
  kbl(caption="Cross-tabulation of age and current grade of student") %>%
  kable_minimal()
```

@@

```{r check-retention-code}
tapply(acs$below_exp_grade,acs[,c("age","current_grade")], mean)
```

@@

```{r retention-by-grade}
acs %>%
  group_by(current_grade) %>%
  summarize(below_grade=mean(below_exp_grade)) %>%
  ggplot(aes(x=current_grade, y=below_grade))+
  geom_col()+
  scale_y_continuous(labels=scales::percent)+
  labs(x="current grade", y="percent clearly behind grade level for age")+
  theme_bw()
```

@@

```{r retention-by-year}
acs %>%
  group_by(year) %>%
  summarize(below_grade=mean(below_exp_grade)) %>%
  ggplot(aes(x=year, y=below_grade))+
  geom_line()+
  geom_point()+
  scale_x_continuous(breaks=c(2010,2014,2018))+
  scale_y_continuous(labels=scales::percent)+
  labs(x="current grade", y="percent clearly behind grade level for age")+
  theme_bw()
```

Overall, there has been some decline over time in the prevalence of grade retention.

Lets look at this by grade and year

```{r retention-by-year-grade}
acs %>%
  group_by(current_grade, year) %>%
  summarize(below_grade=mean(below_exp_grade)) %>%
  ggplot(aes(x=year, y=below_grade, group=current_grade, color=current_grade))+
  geom_line()+
  geom_point()+
  scale_x_continuous(breaks=c(2010,2014,2018))+
  scale_y_continuous(labels=scales::percent)+
  scale_color_viridis_d()+
  labs(x="current grade", y="percent clearly behind grade level for age")+
  theme_bw()
```

The evidence here suggests that the decline is driven by the higher grades where grade retention was higher overall.

# Descriptive statistics

Lets looks at sample characteristics by race.

```{r desc-stats}
tbl_summary <- acs %>%
  group_by(race) %>%
  summarize(n=n(), 
            held_back=round(mean(below_exp_grade), 3)*100,
            mean_age=round(mean(age), 1),
            f_born=round(mean(foreign_born), 3)*100,
            par_fb=round(mean(foreign_born_father | 
                                       foreign_born_mother), 3)*100,
            fam_inc=round(mean(family_income), 0),
            own_home=round(mean(own_home), 3)*100,
            par_hs=round(mean(degree_father!="LHS" | 
                                degree_mother!="LHS"), 3)*100,
            par_ba=round(mean(degree_father=="BA" | 
                                degree_father=="G" | 
                                degree_mother=="BA" | 
                                degree_mother=="G"), 3)*100,
            par_mar=round(mean(parents_married), 3)*100) %>%
  arrange(desc(n))

index_biracial <- which(str_detect(tbl_summary$race, "/"))

tbl_summary %>%
  mutate_if(is.numeric, format, big.mark=",") %>%
  kbl(caption="Descriptive statistics by race. Shading indicates biracial group. Groups ordered by sample size.", align = c("l",rep("r", ncol(tbl_summary)-1))) %>%
  kable_paper() %>%
  row_spec(index_biracial, background = "#D3D3D3")
```

# Model building time

@@

```{r build-models, results='asis'}
model0 <- glm(below_exp_grade~race, data=design_matrix, family=binomial)
model1 <- update(model0, 
                 .~.+y2011+y2012+y2013+y2014+y2015+y2016+y2017+y2018+y2019+
                   current_grade1st+current_grade2nd+current_grade3rd+
                   current_grade4th+current_grade5th+current_grade6th+
                   current_grade7th+current_grade8th+current_grade9th+
                   current_grade10th+current_grade11th+current_grade10th+
                   current_grade11th+current_grade12th+
                   met_non_metro+met_central_city+met_non_central_city+
                   met_city_indeter+
                   s1+s2+s3+s4+s5+s6+s7+s8+s9+s10+s11+s12+s13+s14+s15+s16+s17+
                   s18+s19+s20+s21+s22+s23+s24+s25+s26+s27+s28+s29+s30+s31+s32+
                   s33+s34+s35+s36+s37+s38+s39+s40+s41+s42+s43+s44+s45+s46+s47+
                   s48+s49+s50)
model2 <- update(model1, .~.+foreign_born+foreign_born_mother+
                   foreign_born_father)
model3 <- update(model2, .~.+degree_motherHS+degree_motherAA+degree_motherBA+
                   degree_motherG+degree_fatherHS+degree_fatherAA+
                   degree_fatherBA+degree_fatherG)
model4 <- update(model3, .~.+income_trans+own_home+parents_married)

@@

```{r create-predictor-data}
#dateset for predict
df_pred <- data.frame(race=levels(acs$race),
                      y2011=0, y2012=0, y2013=0, y2014=0, y2015=0, y2016=0,
                      y2017=0, y2018=0, y2019=0,
                      current_grade1st=0, current_grade2nd=0, current_grade3rd=0,
                      current_grade4th=0, current_grade5th=0, current_grade6th=0,
                      current_grade7th=0, current_grade8th=0, current_grade9th=0,
                      current_grade10th=0, current_grade11th=0, current_grade12th=0,
                      met_non_metro=0, met_central_city=0, met_non_central_city=0,
                      met_city_indeter=0,
                      s1=0, s2=0, s3=0, s4=0, s5=0, s6=0, s7=0, s8=0, s9=0, s10=0,
                      s11=0, s12=0, s13=0, s14=0, s15=0, s16=0, s17=0, s18=0, s19=0,
                      s20=0, s21=0, s22=0, s23=0, s24=0, s25=0, s26=0, s27=0, s28=0,
                      s29=0, s30=0, s31=0, s32=0, s33=0, s34=0, s35=0, s36=0, s37=0,
                      s38=0, s39=0, s40=0, s41=0, s42=0, s43=0, s44=0, s45=0, s46=0,
                      s47=0, s48=0, s49=0, s50=0,
                      degree_motherHS=0, degree_motherAA=0, degree_motherBA=0,
                      degree_motherG=0, degree_fatherHS=0, degree_fatherAA=0, 
                      degree_fatherBA=0, degree_fatherG=0,
                      foreign_born=0, foreign_born_father=0, foreign_born_mother=0,
                      income_trans=0, own_home=0, parents_married=0)
```

@@

```{r calc-cond-means}
coef_table <- calculate_cond_means(model0)
```

@@

```{r example-plot}
g <- coef_table %>%
  filter(mrace=="Black/White") %>%
  ggplot(aes(x=term, y=estimate, color=multiracial,
             ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial & mrace=="Black/White"))+
  geom_point()+
  coord_flip()+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="predicted probability of being behind expected grade",
       title="Probability of being behind grade for biracial black/white respondents and monoracial\ncomparison groups",
       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")

g+annotate("text", 1, 0.037, hjust="left",
             label="Grey bands give 83%\nconfidence interval around\nmonoracial estimates")+
  annotate("text", 2, 0.0475, hjust="left",
             label="Red band gives 83%\nconfidence interval around\nmidpoint estimate")+
  annotate("text", 2.5, 0.038, hjust="left",
             label="Actual biracial estimate\nshown with 83% confidence\ninterval")
```

This result is from the baseline model with no control variables, so this would be identical to the result you would get by just calculating raw proportions and their binomial standard errors. The grey bands for the monoracial groups bookmark the range of results we might expect and the red band gives what would be the midpoint value between the groups, adjusting for statistical uncertainty. I can then see where the actual biracial estimate falls within this range. The results here indicate that biracial black/white students have probabilities of grade retention much closer to the lower values of white than black students. However, their probability is still a bit higher than that for whites.

@@

```{r cond-means-model4}
coef_table <- calculate_cond_means(model4)

ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial))+
  geom_point()+
  coord_flip()+
  facet_wrap(~mrace, ncol=3, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="predicted probability of being behind expected grade",
       title="Probability of being behind grade for biracial respondents and monoracial comparison groups",
       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
```

@@

```{r cond-means-model-compare, fig.height=12}
coef_table <- calculate_cond_means(model0) %>%
  mutate(model="crude")
coef_table <- calculate_cond_means(model1) %>%
  mutate(model="fixed effects") %>%
  bind_rows(coef_table)
coef_table <- calculate_cond_means(model2) %>%
  mutate(model="cultural resources") %>%
  bind_rows(coef_table)
coef_table <- calculate_cond_means(model4) %>%
  mutate(model="parent resources") %>%
  bind_rows(coef_table) %>%
  mutate(model=factor(model, levels=c("crude","fixed effects",
                                      "cultural resources",
                                      "parent resources")),
         mrace=factor(mrace, levels=c("Black/White",
                                      "Black/Latino",
                                      "Black/Asian",
                                      "Black/Indigenous",
                                      "White/Indigenous",
                                      "White/Latino",
                                      "White/Asian",
                                      "Latino/Asian",
                                      "Indigenous/Latino",
                                      "Indigenous/Asian")))

ggplot(coef_table, aes(x=term, y=estimate, color=multiracial,
                       ymin=estimate-1.386*se, ymax=estimate+1.386*se))+
  geom_rect(aes(xmax=Inf, xmin=-Inf, ymin=estimate_mid-1.386*se_mid,
                ymax=estimate_mid+1.386*se_mid, fill=multiracial), 
            color=NA, alpha=0.2)+
  geom_linerange(data=subset(coef_table, multiracial))+
  geom_point()+
  coord_flip()+
  facet_grid(mrace~model, scales="free_y")+
  theme_bw()+
  theme(legend.position = "none", panel.grid = element_blank())+
  scale_color_manual(values=c("grey40","red"))+
  scale_fill_manual(values=c("grey40","red"))+
  scale_linetype_manual(values=c(3,1))+
  scale_y_continuous(labels = scales::percent)+
  labs(x=NULL,
       y="predicted probability of being behind expected grade",
       title="Probability of being behind grade for biracial respondents and monoracial comparison groups",
       caption="Non-overlap in confidence bands indicates statistically significant difference at the 5% level")
```

@@

# Sensitivity tests

```{r models-by-age, eval=FALSE}
#run separate models by grade groups
model4a <- update(model4, data=subset(acs, as.numeric(current_grade)<=6))
model4b <- update(model4, data=subset(acs, as.numeric(current_grade)>6 &
                                        as.numeric(current_grade)<=9))
model4c <- update(model4, data=subset(acs, as.numeric(current_grade)>=10))
